---
title: "Reproducible Research: Peer Assessment 1"
author: "Ron Segal"
date: "`r format(Sys.time(), '%B %d %Y')`"
output:
  html_document:
    css: style.css
    theme: "readable"
    highlight: "haddock"
    keep_md: true
---

### Script was created on Friday 17th April 2015

## Initialisation

*Note - Created and run on Ubuntu Linux 14.04, using R Studio Version 0.98.1091*

```{r set_directory, echo=TRUE}
setwd("~/Coursera/RepRes/RepData_PeerAssessment1")
```

### Load required packages

```{r load_packages, echo=TRUE}
#install.packages("broman")
library(broman) # included for its rounding function
library(ggplot2)
library(plyr)
library(grid)
```

## Loading and preprocessing the data

```{r load_data, echo=TRUE}
# Download data if activity.zip doesn't exist
if (!file.exists("activity.csv")) {  
  if (!file.exists("activity.zip")) {
      dataUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
      download.file(dataUrl, destfile="activity.zip", method="wget")
  }
  unzip(zipfile="activity.zip",exdir="./")
}  
```

```{r read_data, echo=TRUE}
# Read in data
activitydata<-read.csv(file="activity.csv", head=TRUE, sep = ",")
```

## What is mean total number of steps taken per day?

```{r sum_days, echo=TRUE}
# Create data frame with sum of steps per day
stepstotal<-ddply(activitydata, c("date"), summarise, stepsum = sum(steps))

qplot(stepstotal$stepsum, geom="histogram", binwidth=800, xlab="Total Steps", main="Histogram of Sum of Steps per Day", fill=I("blue"),alpha=I(.4))

meansteps<-as.integer(mean(stepstotal$stepsum,na.rm=TRUE))
mediansteps<-median(stepstotal$stepsum, na.rm=TRUE)

```

The mean total number of steps per day is: **`r meansteps`**

The median total number of steps per day is: **`r mediansteps`**

## What is the average daily activity pattern?

```{r mean_intervals, echo=TRUE}
# Calculate the mean steps per interval across all days
stepsmean<-ddply(activitydata,c("interval"), summarise, mean = mean(steps,na.rm=TRUE))

ggplot(stepsmean, aes(x=stepsmean$interval, y=stepsmean$mean)) + geom_line(color="blue", stat="identity",binwidth=2) + labs(x="5 Minute Intervals", y="Mean Steps per Interval") + ggtitle("Mean Steps Per Interval over All Days")

```

The 5-minute interval (from `r min(stepsmean$interval)` to `r max(stepsmean$interval)` minutes), on average across all the days in the dataset, that contains the maximum number of steps is: **`r stepsmean[which(stepsmean$mean == max(stepsmean$mean)),1]`** or
`r sprintf("%.1f",stepsmean[which(stepsmean$mean == max(stepsmean$mean)),1] / 300)` hours into the measured period for the day.

## Imputing missing values

```{r total_missing, echo=TRUE}
# Calculate number of rows with missing data
# Only the steps variable column contains missing data
totalmissing<-sum(is.na(activitydata$steps))
```

The total number of missing values in the dataset is: **`r totalmissing`**

Missing steps values are for entire days, so intra day interrpolation wouldn't help.
Consequenlty the simple strategy used to fill in all of the missing values is to replace them with means of corresponding intervals calculated across all days. 

```{r fill_missing, echo=TRUE}
# Create new data set with column containing interval means across all day
newdata<-join(activitydata,stepsmean, by = "interval")
# Replace missing values in step column with corresponding interval means
newdata$steps[is.na(newdata$steps)]<-newdata$mean[is.na(newdata$steps)]

stepstotal<-ddply(newdata, c("date"), summarise, stepsum = sum(steps))
qplot(stepstotal$stepsum, geom="histogram", binwidth=800, xlab="Total Steps", main="Histogram of Sum of Steps per Day - Missing Data Imputed", fill=I("red"),alpha=I(.4))

meansteps<-as.integer(mean(stepstotal$stepsum,na.rm=TRUE))
mediansteps<-as.integer(median(stepstotal$stepsum, na.rm=TRUE))

```

Adjusted for missing data, the mean total number of steps per day is: **`r meansteps`**

Adjusted for missing data, the median total number of steps per day is: **`r mediansteps`**

## Are there differences in activity patterns between weekdays and weekends?

```{r weekday_factor, echo=TRUE} 

# Set time zone to Coordinated Universal Time
Sys.setenv(TZ='UTC')

# Create factor variable
weekDay<-c('Monday','Tuesday','Wednesday','Thursday','Friday')

# Add weekday/weekend 2 level factor column to data frame
newdata$day<-factor((weekdays(as.Date(newdata$date)) %in% weekDay)+1L, levels=1:2, labels=c('weekend','weekday'))
```

*NB - the time zone of the experimental data isn't clear so UTC has been assumed and specifically set. This gives the first day of the data set as:* ***`r weekdays(as.Date(newdata$date[1]))`*** 

```{r weekday_compare, echo=TRUE, fig.width=6, fig.height=6,fig.align='center'}
# Calculate the mean steps per interval across days
daytype<-ddply(newdata, c("interval","day"), summarise, mean = mean(steps))

# The ggplot itself creates panels depending on the factor level in the day variable
ggplot(daytype, aes(x=interval, y=mean)) + geom_line(color="blue", stat="identity",binwidth=2) + labs(x="5 Minute Intervals", y="Mean Steps per Interval") + ggtitle("Mean Steps Per Interval over Weekends vs Weekdays") + facet_wrap(~day,ncol=1) + theme(strip.text.x=element_text(size=12, colour = "brown")) + theme(strip.background = element_rect(fill="pink"))
```
Interpretation: It appears from eyeballing the area under the curves that there is generally more activity at weekends (implication perhaps that the subject has a desk job!)

